Early environmental conditions do not impact behavioural flexibility in an invasive and non-invasive lizard species

### 2024-08-30 ms.qmd
Updates

## Abstract
Behavioural flexibility is expected to be crucial for animals adapting to environmental changes. However, the conditions experienced during early development can profoundly impact behavioural flexibility making it unclear how populations will respond to novel circumstances. Stressful situations faced by the parents can have a direct impact on animals' cognition through the transmission of glucocorticoids - stress-related hormones that affect offspring cognition. At the same time, stressful conditions can influence parental behaviour during nesting and consequently the thermal developmental conditions they experience. Here, we investigated the interactive effects of prenatal corticosterone (CORT) and temperature on behavioural flexibility in two lizard species, _Lampropholis delicata_ and _L. guichenoti_. We manipulated prenatal CORT levels and incubation temperature in a 2x2 factorial design, and then assessed behavioural flexibility through a reversal learning task. We hypothesized prenatal CORT and cold temperatures to impair with lizards' performance in the reversal task. Given _L. delicata_'s success as an invassive species, we expected this species to behave more flexibly and be less affected by early environmental conditions. Contrary to our expectations, behavioural flexibility appears to be robust to prenatal temperature and CORT in both species. The lack of difference in reversal learning between _L. delicata_ and _L. guichenoti_ suggests that novel environments are unlikely to influence flexible behavioural learning and that behavioural flexibility itself is unlikely to explain differences in invasion success between these species.

## Introduction
Behavioural flexibility describes the ability of individuals to adjust behaviour adaptively in response to changes in the internal or external environment [@brown2010behavioral]. It comprises a suite of behaviours, including problem-solving or innovation [@brown2010behavioral], that can be beneficial for adapting to novel circumstances [@szabo2020can; @szulkin2020urban]. Species with greater behavioural flexibility are expected to be more resilient to environmental changes because behaving flexibly is predicted to be crucial for exploiting new resources and facing new challenges [@wright2010behavioral; @chapple2012can; @szabo2020can; @szulkin2020urban]. For example, survival of cane toads in semi-arid regions of Australia depend on adjusting their hydration activities to include daytime hours, rather than relying solely on nighttime [@webb2014behavioural]; while succesful colonisation of new environments by birds is related to higher frequency of foraging innovations [@sol2000behavioural]. Although there is limited understanding of the physiological mechanisms involved in animals' ability to behave flexibly, relative brain size has been shown to be a good predictor of behavioural flexibility in mammals, birds, reptiles and amphibians [@sol2000behavioural; @sol2008brain; @amiel_smart_2011]. This suggests that behavioural flexibility is likely to be underpinned by complex neural mechanisms, and that alterations in these mechanisms may impact animals' ability to respond to novel situations. The brain is considered especially sensitive to environmental changes during the early stages of development [@zhu_prenatal_2004], creating uncertainty about how the conditions faced during early life might impact behavioural flexibility.

The conditions in which organisms develop can be altered by the circumstances faced by their parents. For instance, exposing threespine sticklebacks females to a model predator during ovulation, affected their offspring's tedency to use social cues [@feng2015effect]. Under stressful situations, animal's sensory systems detect and transmit environmental signals to instigate adaptive physiological and behavioural adjustments – responses that are mediated by glucocorticoids (GCs) [@sapolsky_how_2000]. Glucocorticoids can be transmitted to offspring and influence their phenotype and development through transgenerational effects [@Crino_2023]. In addition to the environments experienced by parents, offspring also experience potentially stressful environmental conditions that can interact with or amplify parental effects. For example, human disturbance alters snapping turtles' nest-site choice, ultimately impacting the temperature at which eggs are incubated [@kolbe2002impact]. Because animals' responses to abrupt changes in temperature are mediated by GCs [@Crino_2023], and this response can affect animals' decision-making while nesting, GCs transmission and early therma environment are predicted to interact to shape offspring phenotype. Both GCs and temperature are known to exert potent effects on brain development [@jonson1976effect; @coomber_independent_1997; @zhu_prenatal_2004; @amiel_effects_2017] and cognition [@clark_colour_2014; @lui2017chronic], but little is known about how they may interact to affect behavioural flexibility. Since behaving flexibly is essential for facing novel environments, studying the interactive effect of early environment on behavioural flexibility is crucial if we aim to understand animals' capability to adapt to environmental changes.

Here, we studied the effects of prenatal temperature and corticosterone (CORT) (the main GC in reptiles) [@Crino_2023], on behavioural flexibility in two species of reptiles, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). Both species have similar life-histories and overlapping distributions [@chapple_know_2011; @chapple_biology_2014], but only _Lampropholis delicata_ has been successful in colonising different areas around the globe [@baker1979rainbow; @chapple2013divergent; @chapple2015deliinvLHI]. The differences in invassion success between both skinks may be driven by disparities in behavioural flexibility - a prediction supperted in other invasive species [@sol2000behavioural; @sol2008brain; @wright2010behavioral; @amiel_smart_2011; @chapple2012can]. However, the relative impact of early environments on behavioural flexibility remains unexplored. Early environments are known to affect suites of morphological and behavioural traits differently in each species [@carrasco2023maternal; @kar2023heritability], but we are uncertain about the interactive role of maternal CORT and early temperatures on behavioural flexibility. We manipulated CORT and incubation temperature in both species and then subjected hatchlings to a reversal learning task. We predicted that both cold incubation temperatures and CORT treatment will impair behavioural flexibility [@clark_colour_2014; @lui2017chronic], and the combination of both treatments to have the most detrimental effect on behavioural flexibility. Since invassive species are expected to behave more flexibly [@sol2000behavioural; @sol2008brain; @wright2010behavioral; @amiel_smart_2011; @chapple2012can], we predicted that _L. delicata_, on average, will perform more proficiently in the task compared to _L. guichenoti_ and that behavioural flexibility in _L. delicata_ will be more robust to early environmental conditions.

@@

#### Husbandry  
*Breeding colony* -- We tested juveniles originating from a breeding colony established in the lab since 2019. A total of 270 and 180 adults of *L. delicata* and *L. guichenoti*, respectively, were established in containers (41.5 L x 30.5 W x 21 H cm) with six lizards (2 males and 4 females) per enclosure. Enclosures had non-stick matting, a shelter, and several small water dishes. Water was given daily, and they were fed approx. 40 mid-size crickets (*Acheta domestica*) per enclosure three days a week. Crickets were dusted with calcium weekly and multivitamin and calcium biweekly. Using a heat chord and a heat lamp we created a temperature gradient and kept lights on a 12 h light:12 h dark cycle. UVA/UVB lighting was also provided. Room temperatures were set to 22-24ºC, and warm side of enclosures reached 32ºC.

*Eggs collection and incubation* -- Between mid-October 2022 to the end of February 2023, we provided females with a place to lay eggs by means of small boxes (12.5 L x 8.3 W x 5 H cm) with moist vermiculite inside that were placed in one extreme of the communal enclosures (see above). We checked for the presence of eggs in the boxes three days a week. After collection, we measured the length and width of eggs with a digital caliper to the nearest 0.1 mm and weighed them with a digital scale ± 0.001g error (OHAUS, Model spx123). We also recorded the clutch and egg identities. Eggs were then treated with CORT or vehicle (see CORT and Temperature manipulation below) and were placed in individual cups (80 mL) with moist vermiculite (12 parts water to 4 parts vermiculite). The cups were covered with cling wrap to retain moisture and left in LATWIT 2X5D-R1160 incubators at two different temperatures (see CORT and Temperature manipulation below) until hatching.

Hatchlings – Eggs in the incubator were checked three times a week for hatchlings. After hatchling, we measured juveniles’ SVL and Tail Length (TL) with a ruler to the nearest mm and weighed them with a digital scale ± 0.001g error. We then placed hatchlings in individual enclosures (18.7L x 13.2W x 6.3H cm) and provided them with non-stick matting and a small water dish. During this period, they were sprayed with water every day and received 3-6 small A. domestica crickets three times a week. All care otherwise followed similar protocols to adults (see above).
Two weeks before starting the training phase (see below), lizards were moved to an experimental arena for acclimatation. The arenas were individual medium size (41 L x 29.7 W x 22 H cm) plastic containers with a shelter (9 L x 6 W x 1.5 H cm) on one of the side and a water dish on the other. These new enclosures were placed in two rooms across 7 different racks associated to 7 different CCTV systems (device model DVR-HP210475) that allowed us to record their behaviour during the experiment (see details below). The number of lizards per species and treatment in each rack was counterbalanced to control for any effect of the room or the position of the lizard in the rack. During acclimatation and throughout the experiment, lizards were fed with only one cricket per day dusted with calcium and multivitamin (see protocol below), and water was supplied ad libitum. We provided a temperature gradient by means of a heat cord and heat lamps in a 12 h light: 12 h dark cycle. The temperature of both rooms was set to between 22-24ºC.

@@

To test empirically the effect of early environment we manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (Cold -- 23ºC ± 3ºC or Hot -- 30ºC ± 3ºC) in a 2x2 factorial design (@fig-Methods A). We first allocated eggs into one of two different treatments: in the CORT treatment, eggs were topically supplied with 5µL of CORT dissolved in 100% Ethanol (vehicle) at a final concentration of 10 pg CORT/mL; in the Control treatment eggs received an equal volume of the vehicle. CORT concentration employed in the CORT treatment represents 2 standard deviations above the mean natural concentration obtained in eggs from both species (non-published data). Eggs were then incubated in one of the two previously mentioned temperature regimes (Cold or Hot) until hatching. These temperatures represent the upper and lower limit of the natural incubation temperatures [@qualls2000post; @cheetham2011embryonic]. The number of eggs per clutch assigned to each hormone and temperature treatment were counterbalanced in both species.

@@

In the colour association (Associative task in Fig. @fig-Methods) and reversal task (Reversal task in @fig-Methods B), we used three ramps that were identical in shape and size but different in the colour. The colours of these ramps were green, red, and blue, as previous studies demonstrate that squamates can discriminate between these colours [@Baden_Osorio_2019_Vert_vision]. However, to control for potential colour biases, we assigned one group of lizards within each treatment to associate food in the blue ramp followed by the red ramp in the reversal, while in the other group the order was reversed. As in the last stage of the habituation phase, we placed the cricket inside the wells of the ramps, but then two of them were covered with 3D-printed lids (3D x 0.5H cm) so prey was only accessible in one ramp. The food reward was placed in all three ramps to avoid lizards using prey chemical cues that were released in the closed wells through a series of small holes on the top of the lids. In all trials, the position of the feeders was changed randomly to ensure subjects were using colour rather than spatial cues for the association. Lizards were tested in the association task once a day for 35 days and once a day for 40 days in the reversal.

@@

#### Statistical analyses  
We performed the analyses for each species separately. We fitted two different Bayesian multilevel models using the brm function from the _brms_ package [@burkner2017brms] using an R extension (version 2.8.2) [@R]. We ran four parallel MCMC chains of 3000 iterations for each model, with a warmup period of 1000 iterations. We modelled correct choices [correct (1) or not (0)] as the response variable, and trial, hormone (CORT versus Control), incubation temperature (Cold versus Hot), and their interaction as fixed factors. We included a random intercept and slope (trial) for each lizard in our models. We also incorporated the clutch identity as a random factor. Learning can also be age-dependent (see @noble_age-dependent_2014), and given that lizards incubated at different temperatures hatch at different times we explored the effects of age on learning. Indeed, lizards’ age at the beginning of the experiment ranged from 41 to 148 days old in _L. delicata_ and 48 to 132 in _L. guichenoti_. However, when this variable was included in the models we did not find any significant effect of age (see Supplementary material). As such, we present models without age as a fixed effect. We used the posterior distributions of parameters from these models to test for differences in learning rate between treatments and species. Learning slopes were obtained using the ‘trial’ estimates and its interaction with hormone and temperature treatments. Slope estimates greater than zero were considered as evidence of learning, while those less or equal to zero were not. We calculated the probability that slopes or contrasts between slopes differed from zero (p~mcmc~) using the posterior distributions. We considered statistical significance if p~mcmc~ < 0.05.

#### Ethical note
We collected subjects from a breeding colony kept in the lab since 2019. Both the breeding animals and the ones used in this experiment were provided humane laboratory housing, with thermoregulatory opportunities, light (UV and heat) and moderate levels of humidity (see above for details). Lizards in the colony received food ad libitum while the subjects in the experiment were feed one cricket daily during the experiment. All lizards were handled minimally to avoid stress. Experimental procedures and laboratory housing complied with Australian law and were approved by the Australian National University Animal Experimentation Ethics Committee (A2022_33). Animals born in captivity cannot be re-released into the wild as string biosafety protocols across the states prohibit this. It is also not possible to rehome such a large number of animals for experiments. Therefore, upon completion of experimental trials, lizards were kept in their enclosure until they were humanely euthanized following approved procedures. To euthanise animals, we use an injection of sodium pentobarbital (i.e., Lethabarb) (diluted 1:4 ringers solution) at
60-100mg/kg, injected intrapleuroperitoneally. We monitored the animals to ensure there was no irritation from the agent as indicated by distressed animals. Gently pinching the toes was used to confirm the absence of a response before the animal was disposed of.

@@

Both species learnt across all treatments (@fig-results B, D), but we did not find any difference between treatments or species in the rate of learning. For _L. delicata_, hormone treatment (Control - CORT = `r format_dec(mean(c(deli_ControlCold, deli_ControlHot)) - mean(c(deli_CORTCold, deli_CORTHot)), 3)`, p~mcmc~ `r format_p(pmcmc(c(deli_ControlCold, deli_ControlHot) - c(deli_CORTCold, deli_CORTHot)), 3)`), incubation temperature (Hot - Cold = `r format_dec(mean(c(deli_CORTHot, deli_ControlHot)) - mean(c(deli_CORTCold, deli_ControlCold)),3)`, p~mcmc~ `r format_p(pmcmc(c(deli_CORTHot, deli_ControlHot) - c(deli_CORTCold, deli_ControlCold)),3)`), or the interaction ([(Control-Hot - CORT-Hot) - (Control-Cold - CORT-Cold)] = `r format_dec((mean(deli_ControlHot) - mean(deli_CORTHot)) - (mean(deli_ControlCold) - mean(deli_CORTCold)), 3)`, p~mcmc~ `r format_p(pmcmc((deli_ControlHot - deli_CORTHot) - (deli_ControlCold - deli_CORTCold)), 3)`) did not affect learning (see @fig-results A, B). We similarly found no significant effects of  hormone (Control - CORT = `r format_dec(mean(c(guich_ControlCold, guich_ControlHot)) - mean(c(guich_CORTCold, guich_CORTHot)), 3)`, p~mcmc~ `r format_p(pmcmc(c(guich_ControlCold, guich_ControlHot) - c(guich_CORTCold, guich_CORTHot)), 3)`), incubation temperature (Hot - Cold = `r format_dec(mean(c(guich_CORTHot, guich_ControlHot)) - mean(c(guich_CORTCold, guich_ControlCold)),3)`, p~mcmc~ `r format_p(pmcmc(c(guich_CORTHot, guich_ControlHot) - c(guich_CORTCold, guich_ControlCold)),3)`), or the interaction ([(Control-Hot - CORT-Hot) - (Control-Cold - CORT-Cold)] = `r format_dec((mean(guich_ControlHot) - mean(guich_CORTHot)) - (mean(guich_ControlCold) - mean(guich_CORTCold)), 3)`, p~mcmc~ `r format_p(pmcmc((guich_ControlHot - guich_CORTHot) - (guich_ControlCold - guich_CORTCold)), 3)`) on learning in _L. guichenoti_ (see @fig-results C, D).

@@

Finally, when all individuals from all the treatments were pooled together and species were compared, there were not significant differences in the estimated slopes (mean of learning slope in _L. delicata_ = `r format_dec(mean(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot)), 3)`, 95% CI = [`r format_dec(quantile(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot), 0.025), 3)`, `r format_dec(quantile(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot), 0.975), 3)`]; mean learning of slope in _L. guichenoti_ = `r format_dec(mean(c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot)), 3)`, 95% CI = [`r format_dec(quantile(c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot), 0.025), 3)`, `r format_dec(quantile(c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot), 0.975), 3)`]; _L. delicata_ - _L. guichenoti_ = `r format_dec(mean(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot)) - mean(c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot)), 3)`, p~mcmc~ `r format_p(pmcmc(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot) - c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot)), 3)`).

@@

#### Checking the models plots

Table 1. Estimates of Reversal learning slope for all the different treatments per each task, species, and group. Mean shows the arithmetic means of the estimates obtained from the posteriors of the model, and 95% CI indicates the 95% confidence interval of the mean. All pmcmc tested the hypothesis that the mean equals zero. In bold, those values that are significant (pmcmc <0.05)

@@

>>>>

## Abstract
Behavioural flexibility is expected to be crucial for animals adapting to environmental changes. However, the conditions experienced during early development can profoundly impact behavioural flexibility making it unclear how populations will respond to novel circumstances. Stressful situations faced by the parents can have a direct impact on offspring cognition through the transmission of glucocorticoids - stress-related hormones that affect offspring cognition. At the same time, stressful conditions can influence parental behaviour during nesting and consequently the thermal developmental conditions offspring experience. Here, we investigated the interactive effects of prenatal corticosterone (CORT) and temperature on behavioural flexibility in two lizard species, _Lampropholis delicata_ and _L. guichenoti_. We manipulated prenatal CORT levels and incubation temperature in a 2x2 factorial design, and then assessed behavioural flexibility through a reversal learning task. We hypothesized prenatal CORT and cold temperatures to impair with lizards' performance in the reversal task. Given _L. delicata_'s success as an invassive species, we expected this species to behave more flexibly and be less affected by early environmental conditions. Contrary to our expectations, behavioural flexibility appears to be robust to prenatal temperature and CORT in both species. The lack of difference in reversal learning between _L. delicata_ and _L. guichenoti_ suggests that novel environments are unlikely to influence flexible behavioural learning and that behavioural flexibility itself is unlikely to explain differences in invasion success between these species.

## Introduction
Behavioural flexibility describes the ability of individuals to adjust behaviour adaptively in response to changes in the internal or external environment [@brown2010behavioral]. It comprises a suite of behaviours, including problem-solving or innovation [@brown2010behavioral], that can be beneficial for adapting to novel circumstances [@szabo2020can; @szulkin2020urban]. Species with greater behavioural flexibility are expected to be more resilient to environmental changes because behaving flexibly is predicted to be crucial for exploiting new resources and facing new challenges [@wright2010behavioral; @chapple2012can; @szabo2020can; @szulkin2020urban]. For example, successful colonisation of new environments by birds is related to higher frequency of foraging innovations [@sol2000behavioural]. Although there is limited understanding of the physiological mechanisms involved in animals' ability to behave flexibly, relative brain size has been shown to be a good predictor of behavioural flexibility in mammals, birds, reptiles and amphibians [@sol2000behavioural; @sol2008brain; @amiel_smart_2011]. This suggests that behavioural flexibility is likely to be underpinned by complex neural mechanisms, and that alterations in these mechanisms may impact animals' ability to respond to novel situations. The brain is considered especially sensitive to environmental changes during the early stages of development [@zhu_prenatal_2004], creating uncertainty about how the conditions faced during early life might impact behavioural flexibility.

The conditions in which organisms develop can be altered by the circumstances faced by their parents. For instance, exposing three-spined stickleback (_Gasterosteus aculeatus_) females to a model predator during ovulation, affected their offspring's tedency to use social cues [@feng2015effect]. Under stressful situations, animal's sensory systems detect and transmit environmental signals to instigate adaptive physiological and behavioural adjustments – responses that are mediated by glucocorticoids (GCs) [@sapolsky_how_2000]. Glucocorticoids can be transmitted to offspring and influence their phenotype and development through transgenerational effects [@Crino_2023]. In addition to the environments experienced by parents, offspring also experience potentially stressful environmental conditions that can interact with or amplify parental effects. For example, human disturbance alters snapping turtles' nest-site choice, ultimately impacting the temperature at which eggs are incubated [@kolbe2002impact]. Animals’ responses to abrupt temperature changes are mediated by GCs [@Crino_2023], which can influence animals’ decision-making while nesting [@kolbe2002impact]. As a result, GCs transmission and early thermal environment are expected to interact and shape offspring traits. Both GCs and temperature are known to exert potent effects on brain development [@jonson1976effect; @coomber_independent_1997; @zhu_prenatal_2004; @amiel_effects_2017] and cognition [@clark_colour_2014; @lui2017chronic], but little is known about how they may interact to affect behavioural flexibility. Since behaving flexibly is essential for facing novel environments, studying the interactive effect of early environment on behavioural flexibility is crucial if we aim to understand animals' capability to adapt to environmental changes.

Here, we studied the effects of prenatal temperature and corticosterone (CORT) (the main GC in reptiles) [@Crino_2023], on behavioural flexibility in two species of reptiles, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). Both species have similar life-histories and overlapping distributions [@chapple_know_2011; @chapple_biology_2014], but only _Lampropholis delicata_ has been successful in colonising different areas around the globe [@baker1979rainbow; @chapple2013divergent; @chapple2015deliinvLHI]. Previous studies exploring behavioral differences between the two species have found _L. delicata_ to be more exploratory than _L. guichenoti_ [@chapple_know_2011], but no differences in learning were observed between the skinks in an associative learning task [@bezzina2014does]. Disparities in behavioural flexibility may be driving the differences in invasion success between both skinks, a prediction supported in other invasive species [@sol2000behavioural; @sol2008brain; @wright2010behavioral; @amiel_smart_2011; @chapple2012can]. However, the relative impact of early environments on behavioural flexibility remains unexplored. Early environments are known to affect suites of morphological and behavioural traits differently in each species [@carrasco2023maternal; @kar2023heritability], but uncertainty still exists about the interactive role of maternal CORT and early temperatures on behavioural flexibility. We manipulated CORT and incubation temperature in both species and then subjected hatchlings to a reversal learning task. Reversal learning, in particular, is a widely employed tool to measure behavioral flexibility [@gapp2014early; @hurtubise2017effects], as it assesses an individual's ability to alter a learned behavior in response to environmental changes, providing researchers with a clear indicator of their ability to adjust to new conditions [@brown2010behavioral].

We predicted that both cold incubation temperatures and CORT treatment will impair behavioural flexibility [@clark_colour_2014; @lui2017chronic], and the combination of both treatments to have the most detrimental effect on behavioural flexibility. Since invassive species are expected to behave more flexibly [@sol2000behavioural; @sol2008brain; @wright2010behavioral; @amiel_smart_2011; @chapple2012can], we predicted that _L. delicata_, on average, will perform more proficiently in the task compared to _L. guichenoti_ and that behavioural flexibility in _L. delicata_ will be more robust to early environmental conditions.

@@

#### Husbandry  
*Breeding colony* -- We tested juveniles originating from a breeding colony established in the lab since 2019. A total of 270 and 180 adults of *L. delicata* and *L. guichenoti*, respectively, were established in containers (41.5 L x 30.5 W x 21 H cm) with six lizards (2 males and 4 females) per enclosure. Enclosures had non-stick matting, a shelter, and several small water dishes. Water was given daily, and they were fed approx. 40 mid-size crickets (*Acheta domestica*) per enclosure three days a week. Crickets were dusted with calcium weekly and multivitamin and calcium biweekly. Using a heat chord and a heat lamp we created a temperature gradient and kept lights on a 12 h light:12 h dark cycle. UVA/UVB lighting was also provided. Room temperatures were set to 22-24ºC, and the warm side of enclosures reached 32ºC.

*Eggs collection and incubation* -- Between mid-October 2022 to the end of February 2023, we provided females with a place to lay eggs by means of small boxes (12.5 L x 8.3 W x 5 H cm) with moist vermiculite inside that were placed on one side of the communal enclosures (see above). We checked for the presence of eggs in the boxes three days a week. After collection, we measured the length and width of eggs with a digital caliper to the nearest 0.1 mm and weighed them with a digital scale ± 0.001g error (OHAUS, Model spx123). We also assigned identities to the clutch and each egg. Eggs were then treated with CORT or vehicle (see CORT and Temperature manipulation below) and were placed in individual cups (80 mL) with moist vermiculite (12 parts water to 4 parts vermiculite). The cups were covered with cling wrap to retain moisture and left in LATWIT 2X5D-R1160 incubators at two different temperatures (see CORT and Temperature manipulation below) until hatching.

*Hatchlings* – Eggs in the incubator were checked three times a week for hatchlings. After hatchling, we measured juveniles’ SVL and Tail Length (TL) with a ruler to the nearest mm and weighed them with a digital scale ± 0.001g error. We then placed hatchlings in individual enclosures (18.7L x 13.2W x 6.3H cm) and provided them with non-stick matting, a shelter, and a small water dish. During this period, they were sprayed with water every day and received 3-6 small _A. domestica_ crickets three times a week. All care otherwise followed similar protocols to adults (see above).
Two weeks before starting the training phase (see below), lizards were moved to an experimental arena for acclimatisation. The arenas were individual medium size (41 L x 29.7 W x 22 H cm) plastic containers with a shelter (9 L x 6 W x 1.5 H cm) on one of the side and a water dish on the other. These new enclosures were placed in two rooms across 7 different racks that were monitored by 7 different CCTV systems (device model DVR-HP210475) that allowed us to record their behaviour during the experiment (see details below). Conditions in these rooms where identical to the main room, but the number of lizards per species and treatment in each rack was counterbalanced to control for any potential effect of the room or the position of the lizard on the rack. During acclimatisation and throughout the experiment, lizards were fed with only one cricket per day dusted with calcium and multivitamin (see protocol below), and water was supplied _ad libitum_. We provided a temperature gradient by means of a heat cord and heat lamps in a 12 h light: 12 h dark cycle. The temperature of both rooms was set to between 22-24ºC.

@@

To empirically test the effect of early environment we manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (Cold -- 23ºC ± 3ºC or Hot -- 30ºC ± 3ºC) in a 2x2 factorial design (@fig-Methods A). We first allocated eggs into one of two different treatments: in the CORT treatment, eggs were topically supplied with 5µL of CORT dissolved in 100% Ethanol (vehicle) at a final concentration of 10 pg CORT/mL; in the Control treatment eggs received an equal volume of the vehicle. CORT concentration employed in the CORT treatment represents 2 standard deviations above the mean natural concentration obtained in eggs from both species (non-published data). Doses were determined by referencing published yolk corticosterone concentrations in other oviparous species [@lovern2008effects ; @hanover2019corticosterone] and our own unpublished data. After the hormone treatment, were then incubated in one of the two previously mentioned temperature regimes (Cold or Hot) until hatching. These temperatures represent the upper and lower limit of the natural incubation temperatures [@qualls2000post; @cheetham2011embryonic]. The number of eggs per clutch assigned to each hormone and temperature treatment were counterbalanced in both species.

@@

In the colour association (Associative task in Fig. @fig-Methods) and reversal task (Reversal task in @fig-Methods B), we used three ramps that were identical in shape and size but different in the colour. The colours of these ramps were green, red, and blue, as previous studies demonstrate that squamates can discriminate between these colours [@Baden_Osorio_2019_Vert_vision]. However, to control for potential colour biases, we assigned one group of lizards within each treatment to associate food in the blue ramp followed by the red ramp in the reversal, while in the other group the order was reversed. As in the last stage of the habituation phase, we placed the cricket inside the wells of the ramps, but then two of them were covered with 3D-printed lids (3D x 0.5H cm) so prey was only accessible on one ramp. The food reward was placed on all three ramps to avoid lizards using prey chemical cues that were released in the closed wells through a series of small holes on the top of the lids. In all trials, the position of the feeders was changed randomly to ensure subjects were using colour rather than spatial cues for the association. Lizards were tested in the association task once a day for 35 days and once a day for 40 days in the reversal.

@@

#### Statistical analyses  
We performed the analyses for each species separately. We fitted two different Bayesian multilevel models using the brm function from the _brms_ package [@burkner2017brms] using an R extension (version 2.8.2) [@R]. We ran four parallel MCMC chains of 3000 iterations for each model, with a warmup period of 1000 iterations. We modelled correct choices [correct (1) or not (0)] as the response variable, and trial, hormone (CORT versus Control), incubation temperature (Cold versus Hot), and the three-way interaction between trial, hormone, and temperature. The error structure was modeled using a Bernoulli distribution with a logit link function (family = Bernoulli(link = 'logit')). We included a random intercept and slope (trial) for each lizard in our models. We also incorporated the clutch identity as a random factor. _L. delicata_ lays one clutch per year, while _L. guichenoti_ lays two [@chapple_know_2011; @chapple_biology_2014]. Since eggs were collected during half of the breeding season, clutches likely come from different mothers. By including clutch as a random factor, maternal condition should be accounted for in our analyses. Additionally, differential paternity may be captured by including both clutch and lizard identity in the model. Learning can also be age-dependent (see @noble_age-dependent_2014), and given that lizards incubated at different temperatures hatch at different times we explored the effects of age on learning. Indeed, lizards’ age at the beginning of the experiment ranged from 41 to 148 days old in _L. delicata_ and 48 to 132 in _L. guichenoti_. However, when this variable was included in the models we did not find any significant effect of age (see Supplementary material). As such, we present models without age as a fixed effect. We used the posterior distributions of parameters from these models to test for differences in learning rate between treatments and species. Learning slopes were obtained using the ‘trial’ estimates and its interaction with hormone and temperature treatments. Slope estimates greater than zero were considered as evidence of learning, while those less or equal to zero were not. We calculated the probability that slopes or contrasts between slopes differed from zero (p~mcmc~) using the posterior distributions. We considered statistical significance if p~mcmc~ < 0.05.

#### Ethical note
We collected subjects from a breeding colony kept in the lab since 2019. Both the breeding animals and the ones used in this experiment were provided humane laboratory housing, with thermoregulatory opportunities, light (UV and heat) and moderate levels of humidity (see above for details). Lizards in the colony received food ad libitum while the subjects in the experiment were feed one cricket daily during the experiment. All lizards were handled minimally to avoid stress. Experimental procedures and laboratory housing complied with Australian law and were approved by the Australian National University Animal Experimentation Ethics Committee (A2022_33). Animals born in captivity cannot be re-released into the wild as stringent biosafety protocols across the states prohibit this. It is also not possible to rehome such a large number of animals. Therefore, upon completion of experimental trials, lizards were kept in their enclosure until they were humanely euthanized following approved procedures. To euthanise animals, we use an injection of sodium pentobarbital (i.e., Lethabarb) (diluted 1:4 ringers solution) at 60-100mg/kg, injected intrapleuroperitoneally. We monitored the animals to ensure there was no irritation from the agent as indicated by distressed animals. Gently pinching the toes was used to confirm the absence of a response before the animal was disposed of.  
Crickets were maintained in big communal enclosures in the same room where the big colony is, and at the same temperature and light conditions. They were provided vegetables and water ad libitum, and they were given several egg carton shelters to refuge. All lizards were fed alive crickets except during the experiment, when we used crickets frozen at -21 ºC for 24 hours.

@@

Both species learnt across all treatments (@fig-results B, D), but we did not find any difference between treatments or species in the rate of learning (mean slopes per treatment for both species provided in Table 2 in Supplementary Material). For _L. delicata_, hormone treatment (slopes contrast: Control - CORT = `r format_dec(mean(c(deli_ControlCold, deli_ControlHot)) - mean(c(deli_CORTCold, deli_CORTHot)), 3)`, p~mcmc~ `r format_p(pmcmc(c(deli_ControlCold, deli_ControlHot) - c(deli_CORTCold, deli_CORTHot)), 3)`), incubation temperature (Hot - Cold = `r format_dec(mean(c(deli_CORTHot, deli_ControlHot)) - mean(c(deli_CORTCold, deli_ControlCold)),3)`, p~mcmc~ `r format_p(pmcmc(c(deli_CORTHot, deli_ControlHot) - c(deli_CORTCold, deli_ControlCold)),3)`), or the interaction ([(Control-Hot - CORT-Hot) - (Control-Cold - CORT-Cold)] = `r format_dec((mean(deli_ControlHot) - mean(deli_CORTHot)) - (mean(deli_ControlCold) - mean(deli_CORTCold)), 3)`, p~mcmc~ `r format_p(pmcmc((deli_ControlHot - deli_CORTHot) - (deli_ControlCold - deli_CORTCold)), 3)`) did not affect learning (see @fig-results A, B). We similarly found no significant effects of  hormone (Control - CORT = `r format_dec(mean(c(guich_ControlCold, guich_ControlHot)) - mean(c(guich_CORTCold, guich_CORTHot)), 3)`, p~mcmc~ `r format_p(pmcmc(c(guich_ControlCold, guich_ControlHot) - c(guich_CORTCold, guich_CORTHot)), 3)`), incubation temperature (Hot - Cold = `r format_dec(mean(c(guich_CORTHot, guich_ControlHot)) - mean(c(guich_CORTCold, guich_ControlCold)),3)`, p~mcmc~ `r format_p(pmcmc(c(guich_CORTHot, guich_ControlHot) - c(guich_CORTCold, guich_ControlCold)),3)`), or the interaction ([(Control-Hot - CORT-Hot) - (Control-Cold - CORT-Cold)] = `r format_dec((mean(guich_ControlHot) - mean(guich_CORTHot)) - (mean(guich_ControlCold) - mean(guich_CORTCold)), 3)`, p~mcmc~ `r format_p(pmcmc((guich_ControlHot - guich_CORTHot) - (guich_ControlCold - guich_CORTCold)), 3)`) on learning in _L. guichenoti_ (see @fig-results C, D).

@@

Finally, when all individuals from all the treatments were pooled together and species were compared, there were not significant differences in the estimated slopes (mean of learning slope in _L. delicata_ = `r format_dec(mean(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot)), 3)`, 95% CI = [`r format_dec(quantile(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot), 0.025), 3)`, `r format_dec(quantile(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot), 0.975), 3)`]; mean learning of slope in _L. guichenoti_ = `r format_dec(mean(c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot)), 3)`, 95% CI = [`r format_dec(quantile(c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot), 0.025), 3)`, `r format_dec(quantile(c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot), 0.975), 3)`]; _L. delicata_ - _L. guichenoti_ = `r format_dec(mean(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot)) - mean(c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot)), 3)`, p~mcmc~ `r format_p(pmcmc(c(deli_CORTCold, deli_ControlCold, deli_CORTHot, deli_ControlHot) - c(guich_CORTCold, guich_ControlCold, guich_CORTHot, guich_ControlHot)), 3)`). Figures for both species with the raw data were included in the Supplementary Material.

@@

Table 1. Estimates of Reversal learning slope for all the different treatments per each task, specie, and group. Mean shows the arithmetic means of the estimates obtained from the posteriors of the model, and 95% CI indicates the 95% confidence interval of the mean. All pmcmc tested the hypothesis that the mean equals zero. In bold, those values that are significant (pmcmc <0.05).

@@

Figures with raw data
